### 2009-12-14 docs/marcello-dissertation/thesis/chapter6-data-persistence-component-design-and-architecture.tex
This revision includes the revisions of the Feedback of Chapter 6, with some changes of the Appendix (references).
- Reference to the architecture documented after I removed from my thesis reference. The page brings my name on the top of the page as the author, so just put my name for that reference. It is an online documentation, and therefore, I think it is a better contribution for the DSP future users. This documentation might also be prepared as a Technical paper suggested by Prof. Puder. In this way, this reference can be changed later.

- I had just missed these 2 references about the c

The mongoDB database system was selected, as stated in chapter 5, to store 
the collected data from the SF-BEAMS sensor network. NetBEAMS is based on a 
component-based architecture, where each of the components can play different
roles during their lifecycle. It uses the abstraction of Data Producers and 
Data Consumers: while a component produces data to the system, the latter 
consumes \cite{netbeams2009}. In this way, a DSP Data Component was designed
and implemented to provide the integration between NetBEAM and the mongoDB 
database. This integration is based on the Data Sensor Platform (DSP), the
kernel software stack provided by NetBEAMS, as documented in
\cite{netbeams-dsp-architecture}. Finally, details about the implementation of
this design are located in the Appendix Section
\ref{sec:dsp-data-persistence-implementation}.

@@

As described in \cite{netbeams-dsp-architecture}, the DSP Platform was designed
to support SF-BEAMS. As shown in Figure \ref{fig:dsp-persistence-business-process}, 
the event of data collection starts by the DSP Component deployed on each 
DSP Gateway Node \cite{netbeams-dsp-architecture}, by the event ``Measurements
X Collected''. When this DSP Client component finishes the data extraction
process, a new one starts. First, each of the co-located DSP Client processes 
the collected measurements and creates a DSP Message, embodying the collected 
data from the associated sensor device. Each DSP Message is added into a local
temporary outbound queue that waits to be transmitted. Then, an asynchronous 
data transmission event occurs at a specified rate, packaging all DSP
Messages into the body of a DSP Messages Container in order to be serialized
into XML (shown in Listing \ref{stream:dsp-message-serialized-ysi}) and
transferred to the DSP Server host, a centralized server that receives all the
collected data embodied into each of the DSP Messages.

@@

\section{Requirements Specification}

This section outlines the functionalities and constraints that the developed
solution must perform and follow, respectively, obtained by analyzing the main
data collection scenario described in the previous section and reviewing the
requirements defined by the RTC staff members.

@@

\begin{itemize}
  \item \textbf{Search Measurements by Sensor Properties}: "As a marine
  biologist, I would like to search observations by filtering values of the 
  sensor device's properties such as temperature or salinity, so that I can 
  find specific associated values to the observation.";
  \item \textbf{Search Measurements by Sensor Location}: "As an oceanographer,
  I would like to search observations that took place at the geographic
  coordinates (x latitude, y longitude), so that I can assess the area around
  the given coordinates.";
  \item \textbf{Search Measurements by Data collection date}: "As a marine 
  biologist, I would like to search observations that took place last week, 
  so that I can assess past environmental conditions";
  \item \textbf{Annotate Existing Measurements}: "As a estuarine ecologist,
  I would like to annotate observations from the time the oil spill occurred
  in the San Francisco Bay, so that I can maintain historical evidence of 
  the impact of such event.";
  \item \textbf{Export Data to Spreadsheets}: "As a scientist from RTC, I 
  would like to analyze of the observed data from yesterday using a 
  spreadsheet, so that I can verify measurements using Microsoft Excel.";
  \item \textbf{Export Data to OPeNDAP format}: "As a biologist, I would 
  like to export the collected data produced during this month using the
  OPeNDAP data format, so that I can collaborate with other research groups
  that uses this data format.";
  \item \textbf{Access the data through API, Programming Languages}: "As a
  marine biologist who learned the Python scripting language, I would like to
  write a software that connects directly to the database, so that I can
  implement a customized analysis system for NetBEAMS.";
  \item \textbf{Remove Undesired Measurements}: "As an oceanographer, I
  would like to remove specific observations collected yesterday, 
  so that the research group does not use 'junk data'.".
\end{itemize}

@@

Driven by the requirements defined, the foundation of the integration between
the DSP Data Persistence component and the mongoDB database system is the
specification of the Keys that comprise a set of observations collected from
the sensor devices.

@@

MongoDB is defined as a Document-Oriented Database system, which is based on 
the abstraction of the Key-Value-Pair data model, using a binary version of JSON
\cite{json} as its main data representation. MongoDB does not require the creation 
of a database schema, rather the definition of key descriptions that comprises
a document on the database organized into separate collections of relating
documents. In order to describe the keys of a document, the specifications
detail the use of Data Provenance as guidelines to the description of the
collected data. Therefore, the key-value pairs of any instance of the
collected data must include the following key properties:

@@

After analyzing the DSP Platform and the structure of the DSP Messages in
\cite{netbeams-dsp-architecture}, the values of the collected data must be
categorized into two different sets of data: the ``Key Segment'' and the
``Value Segment''. The key segment provides the necessary metadata 
used to describe the collected data. An instance of the design of the 
keys is shown in Listing \ref{file:mongodb-ysi-data-format}:

\begin{itemize}
  \item \textbf{message\underline{ }id}: Defines the DSP Message ID that
  carried the observation from the sensor device to the DSP Server. This ID
  can be used during network and database inspections in order to track messages
  for the DSP Platform;
  \item \textbf{sensor}: Defines the properties of the sensor device;
    \begin{itemize}[label=\textbullet]
        \item \textbf{ip\underline{ }address} Defines the IP Address of the 
        sensor device that produced the data;
        \item \textbf{location}: Defines the geographical coordinates of the
        sensor device, using the Decimals Degree format \cite{decimal-degrees};
            \begin{itemize}[label=\textbullet]
                \item \textbf{latitude} the GPS latitude coordinate of the
                position of the sensor device;
                \item \textbf{longitude}: the GPS longitude coordinate of the
                position of the sensor device.
            \end{itemize}
    \end{itemize}
  \item \textbf{time}: Defines different time dimensions according to Data 
   Provenance;
    \begin{itemize}[label=\textbullet]
      \item \textbf{valid}: it is extracted from the container's date and time
      of creation, which was collected during the sensor device's reading;
      \item \textbf{transaction}: it is extracted from the DSP message
         container's creation time and it is used to identify when the
         transaction, which transferred the message from the DSP Gateway Client
         to the DSP Server, occurred;
    \end{itemize} 
\end{itemize}

An associated ``Value Segment'' for the aforementioned ``Key Segment'' carries
the observations of a given sensor device. The key \textbf{observation} stores
the values of the collected data. For instance, the observation key of
for the YSI device \cite{YSI-Sonde} may contains all the subsets of 
the properties defined for the device, such as battery, temperature and
salinity, among others. An example of such use of the segments is shown in
Listing \ref{file:mongodb-ysi-data-format}.

Based on this foundation, the design of the DSP Data Persistence component
evolved from a top-down approach.

\section{High-Level System Architecture}

Considering the business analysis described in Section
\ref{sec:business-process-analysis}, the addition of a data persistence layer
to the DSP Platform can be achieved by adding a Data Consumer (DC) as 
described by \cite{netbeams-dsp-architecture}. Based on the 
Requirements specification, this component must be responsible for the 
extraction process that takes the payload of the body of a filtered DSP
Measurement Message and converts it to the key-value data model defined
in the previous section. This process guarantees that the collected data is
inserted to a running mongoDB database system and gives covers the use cases
defined. Figure \ref{fig:NetBEAMS-Persistence-Server-Node-Components}
provides an architectural overview of the DSP Server node with the
loosely-coupled components, highlighting the inclusion of a new DSP Component
called ``DSP Data Persistence''.

@@

\section{DSP Data Persistence: OSGi-DSP Bundle Design}

The proposed Data Persistence component must follow the DSP Component
requirements, which specifies the design-pattern of Data Producer and Consumer
mentioned earlier in this chapter. As seen in the UML Package  Dependency
Diagram in Figure \ref{fig:DSP-Data-Persistence-Packages-Dependency}, a new
DSP Component has dependency on three different packages: the OSGi Framework
and the DSP Platform responsible to integrate a new component into the
micro-kernel, and the mongoDB Java driver API that connects to the mongoDB
server.

@@

The class DSPDataPersistenceActivator is the main class responsible for the
activation of the class DSPDataPersistence, as it extends from the OSGi
class BundleActivator. The BundleActivator maintains references to the classes
BundleContext and ServiceReference, both from the OSGi Framework, in order to
maintain the component managed by the OSGi Framework and register the class
DSPDataPersistece as a service to the OSGi environment. Therefore, the class
DSPDataPersistenceActivator is the main communication interface between the 
DSP Platform and the OSGi Framework, are responsible for the orchestration of
the application lifecycle.

@@

The class DSPDataPersistenceActivator is responsible for starting and stopping,
the services provided by the DSPDataPersistence component, as this behavior is
inherited from the OSGi class BundleActivator from OSGi Framework. Following
the specification of the DSP Architecture \cite{netbeams-dsp-architecture},
Figure
\ref{fig:From-OSGi-Framework-to-DSP-Data-PersistenceActivator-Sequence-Diagram} 
shows a UML Sequence Diagram that depicts the steps of the activation of the
DSP component, and summarized as follows:

@@

\begin{enumerate}
  \item During its activation, the DSP Platform will get the name of the
  DSP Data Persistence component from the configuration artifact config.xml as
  seen in Listing \ref{file:dsp-config.xml};
  \item After being selected based on the configuration priority, the 
  DSPBundle artifact is identified and installed by creating an instance of the
  class DSPDataPersistenceActivator, making a call to the method start();
  \item During the activation, an instance of the class DSPDataPersistence
  is created and registered as an OSGi Service;
  \item Upon registering the DSP Data Persistence component, the DSP Platform
  listens for the event serviceUpdate() and triggers the last operations to
  bootstrap the DSP Component:
   \begin{enumerate}
      \item Initialize the component by calling the method initComponent();
      \item Start the component by calling the method startComponent();
      \item Bootstrap messages by calling the method delivery(). This is where
      both the DSP Measurements Messages and Update Messages are delivered by
      the DSP Broker. This concept is based on the listener of Observer
      Design-Pattern \cite{gof}.
   \end{enumerate}
\end{enumerate}

Any configuration parameter required by a DSP Component is sent during its
bootstrap process through the use of an optional DSP Update Message, as shown
in Listing \ref{file:dsp-data-persistence-bootstrap.xml}. The initialization of
the DSP Component Activator and the DSP Component registration are also shown
in Figure \ref{fig:From-OSGi-Framework-to-DSP-Data-PersistenceActivator-Sequence-Diagram}.

@@

Although the DSP Data Persistence component has been instantiated as of Figure
\ref{fig:From-OSGi-Framework-to-DSP-Data-PersistenceActivator-Sequence-Diagram}, 
the process of delivering messages to the DSP Data Persistence component has yet
to be completed. As defined by the requirements, the DSP Component must execute its
function concurrently, using the configuration parameters during its
bootstrap process. The DSP Platform delivers any bootstrap message defined in
the DSP Platform runtime environment, as explained in the Appendix section
\ref{sec:dsp-persistence-boostrap}. However, to deliver DSP Measurement
Messages, a new filter must be provided. As explained in
\cite{netbeams-dsp-architecture}, the DSP Broker uses the DSP Matcher's rules
to select to Data Consumers configured in a given DSP. For this reason, a rule
that filters any remote measurement message received by running DSP Server is
used, as shown in Listing \ref{file:dsp-matcher-config.xml}. The rule defined
in line 63 express that all messages created by any producer on any DSP Host,
either client or sever, has as delivery target the DSP Data Persistence
component. In this way, when DSP Measurement Messages are delivered, the DSP
Broker delivers the DSP Message to the DSP Data Persistence Component.

@@

\begin{itemize}
  \item A bootstrap DSP Update Message is delivered to the DSP Data Persistence
  component when the DSP Platform is initializing the component. For example, the
  class DSPDataPersistence can initialize the class DspDataFlusher to
  flush temporary messages into the database at the rate of
  TRANSIENT\underline{ }DATA\underline{ }FLUSHER\underline{ }DELAY, defined in 
  seconds. 
  Listing \ref{file:dsp-data-persistence-bootstrap.xml} shows an
  implementation of a bootstrap message for the DSP Data Persistence;
  \item Any DSP Measure Message can be persisted by the DSP Data Persistence
  component. The component will add the message to a temporary memory location
  to decrease I/O on the database, when the DSP Broker delivers a DSP
  MeasureMessage to this component, as depicted by the gate "create
  PersistentMessageUnit Gate". This last gate describes the simple procedure
  to add the message into the transient persistent layer. Listing
  \ref{stream:dsp-message-serialized-ysi}.
\end{itemize}

@@

At this point, the DSP Data Persistence component maintains a set of DSP
Messages in a local cache during the specific ``flush rate'' specified during
the bootstrap message. When the class DSPDataFlusher finds any number of
instances of PersistentMessageUnit in cache, it starts the process of flushing
them into the mongoDB database, described in the following section.

@@

\begin{itemize}
  \item \textbf{TIMED\underline{ }WAITING}: the initial state of a Java thread
  after its initialization, waiting to be on the ``runnable'' state. This
  period of time is defined by the ``flush rate''  specified in the bootstrap
  message of the DSP Data Persistence component;
  \item \textbf{RUNNABLE}: the class DSPDataFlusher contacts the class
  TransientPersistenceLayer, in order to verify if there are any instance of
  the class PersistentMessageUnit waiting to be transmitted from the temporary
  cache to the Database service. If the cache is not empty, the DSPDataFlusher 
  acquires the lock over the TransientPersistenceLayer and retrieves the
  existing set of PersistentMessageUnit to be sent to mongoDB. Then, it sends
  all of the PersistentMessageUnit classes to the Singleton Service class 
  DSPMongoCRUDService, which is responsible to extract the collected data from
  the DSP Message's Body and convert it to the Key-Value model specified in
  Section \ref{sec:dsp-persistence-data-model}.
\end{itemize}

The main participating classes from the DSPMongoCRUDService are shown in the
UML Class Diagram \ref{fig:DSP-Data-Persistence-Mongo-Classes}. It keeps references
to all types of sensor devices such as the SondeDataContainer and the
MouseActionsContainer. Furthermore, it is responsible to implement the CRUD
methods to modify the mongoDB database state. The class Mongo is used to acquire
an instance of the connection with mongoDB server and provide Factory Methods 
\cite{gof} to create instances of the classes DB, DBCollection and
BasicDBObject. The class DB is used as a reference to a database namespace, 
such as ``netbeams'', the one suggested and implemented by this work. Similarly,
class DBCollection offers an API to manipulate similar documents via the
abstraction of a collection of items. Finally, the class BasicDBObject is used
to represent the keys and values to be persisted in the database.

@@

In order to build the document key and value segments, as described in
section \ref{sec:dsp-persistence-data-model}, the reference to the DSP message
is used, as the value key is built by extracting the DSP Message Content from
the body of the DSP Message. Similarly, the key is built by gathering
information from the DSP Message and the PersistenceMessageUnit class. Each of
the properties collected from the YSI Sonde sensor device is collected and
associated with the keys defined for the Value Segment.

\begin{figure}[!h]
  \centering
  \includegraphics[scale=0.6]{../diagrams/From-Creating-Value-Segment-Sequence}
  \caption{UML Sequence diagram showing the creation of the value document segment}
  \label{fig:From-Creating-Value-Segment-Sequence}
\end{figure}

@@

Once the Value Segment is created, the Key Segment is prepared as described in
the UML Sequence Diagram in Figure
\ref{fig:From-Creating-Key-Segment-Sequence}.

\begin{figure}[!h]
  \centering
  \includegraphics[scale=0.65]{../diagrams/From-Creating-Key-Segment-Sequence}
  \caption{UML Sequence diagram showing the creation of the key document segment}
  \label{fig:From-Creating-Key-Segment-Sequence}
\end{figure}

To Summarize, the DSP Data Persistence Component is added into the DSP Platform
by introducing a new component that provides service to the transfer any
collected data from sensor devices to mongoDB, the Key-Value Pair database
technology. The general dependencies of the system can be seen in UML
Class Diagram depicted by Figure \ref{fig:DSP-Data-Persistence-Classes}, that
expands the UML Package Dependency Diagram shown in Figure
\ref{fig:DSP-Data-Persistence-Packages-Dependency}.

@@

\section{DSP Data Persistence Deployment}

@@

The deployment of the DSP Data Persistence follows the deployment specifications
of a DSP Component over an OSGi Framework. For the development of this work,
the Knopflerfish OSGi Container was used. The details of implementation are described
in section \ref{sec:dsp-data-persistence-implementation}. The deployment of the 
mongoDB can be accomplished in different approaches, summarized as follows:

@@

Given that mongoDB provides support for Database Partitioning by using
the concept of Shards, the implementation of a Data-Centric approach can be done
by deploying different mongoDB instances in a distributed cluster, where each
node can hold different collections of data. mongoDB provides Horizontal
Partition with automatic distribution based on the ranges of a given key of
the document. For NetBEAMS collected data, a good candidate for partitioning is
the time when the data was created, and separate each shard in groups of month
of the year.

@@

The design of the DSP Data Persistence uses the DSP Component specification
defined by the DSP Platform. First, the analysis of the business process
helped to understand the requirements of the system. The data model design
followed the specifications defined in chapter 5. Each of the layers of the
system was described using UML Diagrams, having partial details regarding the
OSGi implementation. The next chapter entails the experiments conducted to
evaluate the designs discussed in this chapter, and walks through the
experiments conducted to verify the implementation of using mongoDB.

>>>>

The mongoDB database system was selected, as stated in Chapter 5, to store 
the collected data from the SF-BEAMS sensor network. NetBEAMS is based on a 
component-based architecture, where each of the components can play different
roles during their lifecycle. It uses the abstraction of Data Producers and 
Data Consumers: while a component produces data to the system, the latter 
consumes \cite{netbeams2009}. In this way, a DSP Data Component was designed
and implemented to provide the integration between NetBEAMS and the mongoDB 
database. This integration is based on the Data Sensor Platform (DSP), the
kernel software stack provided by NetBEAMS, as documented in
\cite{netbeams-dsp-architecture}.

This chapter follows the steps taken in Software Engineering guidelines to
analyze and design a software component using Object-Oriented Analysis and 
Design \cite{oop} using the techniques from the Agile and XP Scrum
\cite{agile-scrum} processes. First, Section \ref{sec:business-process-analysis} gives an
overview of the problem by analyzing a basic scenario of the data collection process used in NetBEAMS.
Then, the Section \ref{sec:requirements-specs} describes the requirements for
the design of the new component, followed by the specifications of the data
model used for the mongoDB database in Section
\ref{sec:dsp-persistence-data-model}. Lastly, the high-level architecture
of Section \ref{sec:system-architecture} gives the foundation of the
DSP Data Persistence component design of Section \ref{sec:dsp-data-comp-design}.

Details about the implementation of this design are located in the
Appendix Section \ref{sec:dsp-data-persistence-implementation}.

@@

As described in \cite{netbeams-dsp-architecture}, the DSP Platform was designed
to support SF-BEAMS. As shown in Figure \ref{fig:dsp-persistence-business-process}, 
the event of data collection starts by the DSP Component deployed on each 
DSP Gateway Node \cite{netbeams-dsp-architecture}, by the event ``Measurements
X Collected''. When this DSP Client component finishes the data extraction
process, a new one starts. First, each of the co-located DSP Client processes 
the collected measurements and creates a DSP Message, embodying the collected 
data from the associated sensor device. Each DSP Message is added into a local
temporary outbound queue that waits to be transmitted. Then, an asynchronous 
data transmission event occurs at a specified rate, packaging all DSP
Messages into the body of a DSP Messages Container in order to be serialized
into XML (shown in Listing \ref{stream:dsp-message-serialized-ysi} on page
\pageref{stream:dsp-message-serialized-ysi}) and transferred to the DSP Server
host, a centralized server that receives all the collected data embodied into each of the DSP Messages.

@@

\section{Requirements Specification}
\label{sec:requirements-specs}
This section outlines the functionalities and constraints that the developed
solution must perform and follow, respectively, obtained by analyzing the main
data collection scenario described in the previous section and reviewing the
requirements defined by the RTC staff members.

@@

\begin{itemize}
  \item \textbf{Search Measurements by Sensor Properties}: "As a marine
  biologist, I would like to search observations by filtering values of the 
  sensor device's properties such as temperature or salinity, so that I can 
  find specific associated values to the observation.";
  \item \textbf{Search Measurements by Sensor Location}: "As an oceanographer,
  I would like to search observations that took place at the geographic
  coordinates (x latitude, y longitude), so that I can assess the area around
  the given coordinates.";
  \item \textbf{Search Measurements by Data collection date}: "As a marine 
  biologist, I would like to search observations that took place last week, 
  so that I can assess past environmental conditions";
  \item \textbf{Annotate Existing Measurements}: "As a estuarine ecologist,
  I would like to annotate observations from the time the oil spill occurred
  in the San Francisco Bay, so that I can maintain historical evidence of 
  the impact of such event.";
  \item \textbf{Export Data to Spreadsheets}: "As a scientist from RTC, I 
  would like to analyze of the observed data from yesterday using a 
  spreadsheet, so that I can verify measurements using Microsoft Excel.";
  \item \textbf{Export Data to OPeNDAP format}: "As a biologist, I would 
  like to export the collected data produced during this month using the
  OPeNDAP data format, so that I can collaborate with other research groups
  that use this data format.";
  \item \textbf{Access the data through API, Programming Languages}: "As a
  marine biologist who learned the Python scripting language, I would like to
  write a software that connects directly to the database, so that I can
  implement a customized analysis system for NetBEAMS.";
  \item \textbf{Remove Undesired Measurements}: "As an oceanographer, I
  would like to remove specific observations collected yesterday, 
  so that the research group does not use 'junk data'.".
\end{itemize}

@@

Driven by the requirements defined, the definition of the keys of the key-value
pairs model, used to describe the collected data from the sensor devices and
their associated metadata, is the foundation for the integration between the
DSP Data Persistence component and the mongoDB database. Next section details
the data model design that specifies those keys.

@@

MongoDB is defined as a Document-Oriented Database system, which is based on 
the abstraction of the Key-Value-Pair data model, using a binary version of JSON
\cite{json} as its main data representation. MongoDB does not require the creation 
of a database schema, rather the definition of keys that comprise a document on
the database organized into separate collections of relating documents. In
order to describe the keys of a document, the guidelines provided by the Data
Provenance taxonomy was used. The key-value pairs of any instance of the
collected data must include the following key properties:

@@

After analyzing the DSP Message Structure, as well as the YSI Sonde Data
Handler of the DSP Platform architecture documentation
\cite{netbeams-dsp-architecture}, the design of the document developed to
identify both the data and metadata of the collected data is organized as
follows:

\begin{itemize}
  \item \textbf{message\underline{ }id}: Defines the DSP Message ID that
  carried the observation from the sensor device to the DSP Server. This ID
  can be used during network and database inspections in order to track messages
  for the DSP Platform;
  \item \textbf{sensor}: Defines the properties of the sensor device;
    \begin{itemize}[label=\textbullet]
        \item \textbf{ip\underline{ }address} Defines the IP Address of the 
        sensor device that produced the data;
        \item \textbf{location}: Defines the geographical coordinates of the
        sensor device, using the Decimals Degree format \cite{decimal-degrees};
            \begin{itemize}[label=\textbullet]
                \item \textbf{latitude} the GPS latitude coordinate of the
                position of the sensor device;
                \item \textbf{longitude}: the GPS longitude coordinate of the
                position of the sensor device.
            \end{itemize}
    \end{itemize}
  \item \textbf{time}: Defines different time dimensions according to Data 
   Provenance;
    \begin{itemize}[label=\textbullet]
      \item \textbf{valid}: it is extracted from the container's date and time
      of creation, which was collected during the sensor device's reading;
      \item \textbf{transaction}: it is extracted from the DSP message
         container's creation time and it is used to identify when the
         transaction, which transferred the message from the DSP Gateway Client
         to the DSP Server, occurred;
    \end{itemize} 
    \item \textbf{observation}: carries the observations of a given sensor
    device. For example, the observation key of for the YSI device
    \cite{YSI-Sonde} may contains all the subsets of the properties defined
    for the device, such as battery, temperature and salinity, among others.
\end{itemize}

MongoDB uses JSON data structure \cite{json} to represent a document, as the
example shown in Figure \ref{fig:kvp-json-viewer} on page
\pageref{fig:kvp-json-viewer}. Any number of keys can be in the root branch of
the tree, as well as in sub branches of the document structure. The values of
the tree are concentrated in the leaves of the tree. Listing
\ref{file:mongodb-ysi-data-format}, on page
\pageref{file:mongodb-ysi-data-format}, shows the instance of this design on
mongoDB database. Based on this foundation, the design of the DSP Data
Persistence component evolved from a top-down approach to meet this data model
design.

\begin{figure}[!h]
  \centering
  \includegraphics[scale=0.65]{../diagrams/kvp-json-viewer}
  \caption{Data Design using JSON Data Structure}
  \label{fig:kvp-json-viewer}
\end{figure}

It is important to note that mongoDB does not use any structure language, like
the use of the SQL in relational databases, to access the persisted data. In
order to access the values of the documents, the ``dot notation''
\cite{db-mongo-data-access} is used to reach each of the tree nodes. For
example, the use of ``observation.WaterTemperature'' and
``sensor.location.latitude'' give the values of the data ``water temperature'' and
the metadata ``latitude'' of the collected data, respectively.

\section{High-Level System Architecture}
\label{sec:system-architecture}

Considering the business analysis described in Section
\ref{sec:business-process-analysis}, the addition of a data persistence layer
to the DSP Platform can be achieved by adding a Data Consumer (DC) as 
described by \cite{netbeams-dsp-architecture}. Based on the 
Requirements specification, this component must be responsible for the 
extraction process that takes the payload of the body of a filtered DSP
Measurement Message and converts it to the key-value data model defined
in the previous section. This process guarantees that the collected data is
inserted to a running mongoDB database system and covers the use cases
defined. Figure \ref{fig:NetBEAMS-Persistence-Server-Node-Components}
provides an architectural overview of the DSP Server node with the
loosely-coupled components, highlighting the inclusion of a new DSP Component
called ``DSP Data Persistence''.

@@

\section{DSP Data Persistence Component Design}
\label{sec:dsp-data-comp-design}

The proposed Data Persistence component must follow the DSP Component
requirements, which specifies the design-pattern of Data Producer and Consumer
mentioned earlier in this chapter. As seen in the UML Package  Dependency
Diagram in Figure \ref{fig:DSP-Data-Persistence-Packages-Dependency}, a new
DSP Component has dependencies on three different packages: the OSGi Framework
and the DSP Platform responsible to integrate a new component into the
micro-kernel, and the mongoDB Java driver API that connects to the mongoDB
server.

@@

The class DSPDataPersistenceActivator is the main class responsible for the
activation of the class DSPDataPersistence, as it extends from the OSGi
class BundleActivator. The BundleActivator maintains references to the classes
BundleContext and ServiceReference, both from the OSGi Framework, in order to
maintain the component managed by the OSGi Framework and register the class
DSPDataPersistence as a service to the OSGi environment. Therefore, the class
DSPDataPersistenceActivator is the main communication interface between the 
DSP Platform and the OSGi Framework, are responsible for the orchestration of
the application lifecycle.

@@

The class DSPDataPersistenceActivator is responsible for starting and stopping,
the services provided by the DSPDataPersistence component, as this behavior is
inherited from the OSGi class BundleActivator from the OSGi Framework. Following
the specification of the DSP Architecture \cite{netbeams-dsp-architecture},
Figure
\ref{fig:From-OSGi-Framework-to-DSP-Data-PersistenceActivator-Sequence-Diagram} 
shows a UML Sequence Diagram that depicts the steps of the activation of the
DSP component, and summarized as follows:

@@

\begin{enumerate}
  \item During its activation, the DSP Platform will get the name of the
  DSP Data Persistence component from the configuration artifact config.xml as
  seen in Listing \ref{file:dsp-config.xml} on page
  \pageref{file:dsp-config.xml};
  \item After being selected based on the configuration priority, the 
  DSPBundle artifact is identified and installed by creating an instance of the
  class DSPDataPersistenceActivator, making a call to the method start();
  \item During the activation, an instance of the class DSPDataPersistence
  is created and registered as an OSGi Service;
  \item Upon registering the DSP Data Persistence component, the DSP Platform
  listens for the event serviceUpdate() and triggers the last operations to
  bootstrap the DSP Component:
   \begin{enumerate}
      \item Initialize the component by calling the method initComponent();
      \item Start the component by calling the method startComponent();
      \item Bootstrap messages by calling the method delivery(). This is where
      both the DSP Measurements Messages and Update Messages are delivered by
      the DSP Broker. This concept is based on the listener of Observer
      Design-Pattern \cite{gof}.
   \end{enumerate}
\end{enumerate}

Any configuration parameter required by a DSP Component is sent during its
bootstrap process through the use of an optional DSP Update Message, as shown
in Listing \ref{file:dsp-data-persistence-bootstrap.xml} on page
\pageref{file:dsp-data-persistence-bootstrap.xml}. The initialization of the
DSP Component Activator and the DSP Component registration are also shown in
Figure
\ref{fig:From-OSGi-Framework-to-DSP-Data-PersistenceActivator-Sequence-Diagram}.

@@

Although the DSP Data Persistence component has been instantiated as of Figure
\ref{fig:From-OSGi-Framework-to-DSP-Data-PersistenceActivator-Sequence-Diagram}, 
the process of delivering messages to the DSP Data Persistence component has yet
to be completed. As defined by the requirements, the DSP Component must execute its
function concurrently, using the configuration parameters during its
bootstrap process. The DSP Platform delivers any bootstrap message defined in
the DSP Platform runtime environment, as explained in the Appendix section
\ref{sec:dsp-persistence-boostrap}. However, to deliver DSP Measurement
Messages, a new filter must be provided. As explained in
\cite{netbeams-dsp-architecture}, the DSP Broker uses the DSP Matcher's rules
to select to Data Consumers configured in a given DSP. For this reason, a rule
that filters any remote measurement message received by running DSP Server is
used, as shown in Listing \ref{file:dsp-matcher-config.xml} on page
\pageref{file:dsp-matcher-config.xml}. The rule defined in line 63 express
that all messages created by any producer on any DSP Host, either client or
sever, has as delivery target the DSP Data Persistence component. In this way,
when DSP Measurement Messages are delivered, the DSP Broker delivers the DSP
Message to the DSP Data Persistence Component.

@@

\begin{itemize}
  \item A bootstrap DSP Update Message is delivered to the DSP Data Persistence
  component when the DSP Platform is initializing the component. For example, the
  class DSPDataPersistence can initialize the class DspDataFlusher to
  flush temporary messages into the database at the rate of
  TRANSIENT\underline{ }DATA\underline{ }FLUSHER\underline{ }DELAY, defined in 
  seconds. 
  Listing \ref{file:dsp-data-persistence-bootstrap.xml}, on page
  \pageref{file:dsp-data-persistence-bootstrap.xml}, shows an implementation of
  a bootstrap message for the DSP Data Persistence;
  \item Any DSP Measure Message can be persisted by the DSP Data Persistence
  component. The component will add the message to a temporary memory location
  to decrease I/O on the database, when the DSP Broker delivers a DSP
  MeasureMessage to this component, as depicted by the UML Gate element
  \cite{uml-seq-gate} "create PersistentMessageUnit Gate". This last Gate
  describes the simple procedure to add the message into the transient
  persistent layer, as shown in Listing
  \ref{stream:dsp-message-serialized-ysi} on page
  \pageref{stream:dsp-message-serialized-ysi}.
\end{itemize}

@@

At this point, the DSP Data Persistence component maintains a set of DSP
Messages in a local cache during the specific ``flush rate'' specified in
the bootstrap message parameter. When the class DSPDataFlusher finds any number
of instances of PersistentMessageUnit in cache, it starts the process of
flushing them into the mongoDB database, described in the following section.

@@

\begin{itemize}
  \item \textbf{Waiting}: the initial state of a Java thread
  after its initialization, waiting to be on the ``runnable'' state. This
  period of time is defined by the ``flush rate''  specified in the bootstrap
  message of the DSP Data Persistence component;
  \item \textbf{Running}: the class DSPDataFlusher contacts the class
  TransientPersistenceLayer, in order to verify if there are any instance of
  the class PersistentMessageUnit waiting to be transmitted from the temporary
  cache to the Database service. If the cache is not empty, the DSPDataFlusher 
  acquires the lock over the TransientPersistenceLayer and retrieves the
  existing set of PersistentMessageUnit to be sent to mongoDB. Then, it sends
  all of the PersistentMessageUnit classes to the Singleton Service class 
  DSPMongoCRUDService, which is responsible to extract the collected data from
  the DSP Message's Body and convert it to the Key-Value model specified in
  Section \ref{sec:dsp-persistence-data-model}.
\end{itemize}

The main participating classes from the DSPMongoCRUDService are shown in the
UML Class Diagram \ref{fig:DSP-Data-Persistence-Mongo-Classes}. It keeps references
to all types of sensor devices. Furthermore, it is responsible to implement the
CRUD methods to modify the mongoDB database state. The class Mongo is used to acquire
an instance of the connection with mongoDB server and provide Factory Methods 
\cite{gof} to create instances of the classes DB, DBCollection and
BasicDBObject. The class DB is used as a reference to a database namespace, 
such as ``netbeams'', the one suggested and implemented by this work. Similarly,
class DBCollection offers an API to manipulate similar documents via the
abstraction of a collection of items. Finally, the class BasicDBObject is used
to represent the keys and values to be persisted in the database.

@@

In order to build the mongoDB document described in section
\ref{sec:dsp-persistence-data-model}, the reference to the DSP Message is
used to extracted the metadata, while the reference to the DSP Message Content 
from the body of the DSP Message is used to extract the collected data from
sensors.

\begin{figure}[!h]
  \centering
  \includegraphics[scale=0.6]{../diagrams/From-Creating-Value-Segment-Sequence}
  \caption{UML Sequence diagram showing the creation of the keys for the
  collected data}
  \label{fig:From-Creating-Value-Segment-Sequence}
\end{figure}

@@

After the keys for the collected data are created, the keys for the metadata
are captured as shown in the UML Sequence Diagram in Figure
\ref{fig:From-Creating-Key-Segment-Sequence}.

\begin{figure}[!h]
  \centering
  \includegraphics[scale=0.65]{../diagrams/From-Creating-Key-Segment-Sequence}
  \caption{UML Sequence diagram showing the creation of the keys for the
  metadata}
  \label{fig:From-Creating-Key-Segment-Sequence}
\end{figure}

To summarize, the DSP Data Persistence Component is added into the DSP Platform
by introducing a new component that provides service to the transfer any
collected data from sensor devices to mongoDB, the Key-Value Pair database
technology. The general dependencies of the system can be seen in UML
Class Diagram depicted by Figure \ref{fig:DSP-Data-Persistence-Classes}, that
expands the UML Package Dependency Diagram shown in Figure
\ref{fig:DSP-Data-Persistence-Packages-Dependency}.

@@

\subsection{DSP Data Persistence Deployment}

@@

The deployment of the DSP Data Persistence follows the deployment specifications
of a DSP Component over an OSGi Framework. For the development of this work,
the Knopflerfish OSGi Container \cite{knopflerfish} was used. The details of the
implementation are described in Section
\ref{sec:dsp-data-persistence-implementation}. The deployment of the mongoDB
can be accomplished in different approaches, summarized as follows:

@@

Given that mongoDB provides support for Database Partitioning by using
the concept of Shards, the implementation of a Data-Centric approach can be done
by deploying different mongoDB instances in a distributed cluster, where each
node can hold different collections of data. mongoDB provides Horizontal
Partition with automatic distribution based on the ranges of a given key of
the document. For NetBEAMS collected data, a good candidate for partitioning is
the time when the data was created, and separate each shard in groups of months
of the year.

@@

The design of the DSP Data Persistence uses the DSP Component specification
defined by the DSP Platform. First, the analysis of the business process
helped to understand the requirements of the system. The data model design
followed the specifications defined in Chapter 5. Each of the layers of the
system was described using UML Diagrams, having partial details regarding the
OSGi implementation. The next chapter entails the experiments conducted to
evaluate the designs discussed in this chapter, and walks through the
experiments conducted to verify the implementation of using mongoDB.
